---
title: "Clustering with STARRS"
author: "Daphné Giorgi, Antoine Godichon-Baggioni, Stéphane Robin"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    toc: true
    toc_depth: 3
  bookdown::pdf_document2:
    toc: yes
    toc_depth: 3
    number_sections: yes
package: STARRS
bibliography: STARRS.bib
vignette: >
  %\VignetteIndexEntry{Clustering with STARRS}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(STARRS)
set.seed(1)
```

# Introduction

Grouping observations into homogeneous groups is a recurrent task in statistical data analysis. 
Unsupervised classification methods (also called clustering methods) are typically divided into two main classes: distance-based methods, and model-based methods. 
The most representative examples in each category are, respectively, the $k$-means (@McQ67) algorithm and the Gaussian mixture model (@MaP00).
The STARRS package propose robust counterparts of each of them, namely the $k$-medians algorithm (@CCM10) and a robust version of the EM algorithm for Gaussian mixtures (@godichon2024robust).

# Model-based clustering: A first example

## Synthetic data set

A sample of $n=1500$ drawn from a multivariate Gaussian mixture model with $K=3$ components (with equal proportions) contaminated with outliers with student distribution can be obtained with the  function \texttt{rGMMcontStudent}, as follows.

```{r simul, eval=TRUE}
# Dimensions
K <- 3; nk <- rep(500, K); d <- 5
# Gaussian parameters
mu <- rbind(mu1=rep(0, d), mu2=rep(3, d), mu3=rep(-3, d))
sigma <- array(dim=c(K, d, d))
sigma[1, , ] <- 2*diag(d); sigma[2, , ] <- diag(1:d); sigma[3, , ] <- diag(1/(1:d))
# Student contamination
fraction <- 0.3
dfT <- 1; dfT <- 1; muT <- mu; sigmaT <- sigma
# Simulation
GMMcontStudent <- rGMMcontStudent(nk=nk, muG=mu, sigmaG=sigma,
                                  delta=fraction,
                                  dfT=dfT, muT=muT, sigmaT=sigmaT)
print(head(STARRS::GMMcontStudent$X))
print(head(STARRS::GMMcontStudent$Z))
print(head(STARRS::GMMcontStudent$C))
```

## Robust clustering with STARRS

The robust version of the EM algorithm proposed by \cite{GBR24} for the inference of Gaussian mixture models is implemented in the function \texttt{multipleRobustMM}.

```{r rgmm, eval=TRUE}
robustGMMOutput <- multipleRobustMM(GMMcontStudent$X, nclust=1:5)
print(robustGMMOutput$bestresult$centers)
```

The number of components can be chosen with the BIC or ICL criterion.

```{r selection, eval=TRUE}
plot(robustGMMOutput, graph='BIC')
plot(robustGMMOutput, graph='ICL')
```

The classification of the observations (in the first two dimensions of PCA) can be visualized with the \texttt{plot} function of the package, together with confidence of the classification uncertainty.

```{r uncertainty, eval=TRUE}
plot(robustGMMOutput, graph='Two_Dim_Uncertainty')
```

# Robustness study

## Distance-based methods

The first element of the list of datasets \texttt{homoGMMStudentCont_K3_d5_n1500} consists of an uncontaminated sample of $n=1500$ observations drawn from a multivariate Gaussian mixture with $K=3$ components, with equal proportions and variance matrices all equal to twice the identity. The next datasets from this list consists of the same data set contaminated with an increasing fraction $f$ of outliers: $f = 0, 10\%, \dots 50\%$ drawn from a multivariate student mixture. 

Both the regular $k$-means algorithm and the $k$-medians algorithm can be applied to each element of this list.

```{r kmeans, eval=TRUE}
K <- 3; 
# k-means
kmeans <- lapply(1:length(homoGMMStudentCont_K3_d5_n1500), function(f){
  kmeans(homoGMMStudentCont_K3_d5_n1500[[f]]$X, centers=K, nstart=20, iter.max=100)})
# k-medians
kmedians <- lapply(1:length(homoGMMStudentCont_K3_d5_n1500), function(f){
  kmedians(homoGMMStudentCont_K3_d5_n1500[[f]]$X)})
```

The next plot illustrate the effect of contamination on the classification accuracy as measure by the adjusted Rand index (ARI).

```{r kmeansAri, eval=TRUE}
# Classification accuracy
fractionList <- (0:5)/10; library(mclust)
plot(fractionList, sapply(1:length(fractionList), function(ff){
  adjustedRandIndex(heteroGMMStudentCont_K3_d5_n1500[[ff]]$Z, kmeans[[ff]]$cluster)
  }), type='b', pch=20, xlab='fraction of outliers', ylab='ARI')
points(fractionList, sapply(1:length(fractionList), function(ff){
  adjustedRandIndex(heteroGMMStudentCont_K3_d5_n1500[[ff]]$Z, kmedians[[ff]]$bestresult$cluster)
  }), type='b', pch=20, col='red')
```

## Distance-based methods

The first element of the list of datasets \texttt{heteroGMMStudentCont_K3_d5_n1500} has been contructed is the same way as the \texttt{homoGMMStudentCont_K3_d5_n1500} except that the components of the multivariate Gaussian mixture have different variance matrices.

Again, both the regular EM algorithm (as implemente in the \texttt{mclust} package: @SFM16) and its robust version can be applied to each element of this list.

```{r gmm, eval=TRUE}
# Regular EM
gmm <- lapply(1:length(heteroGMMStudentCont_K3_d5_n1500), function(f){
  Mclust(heteroGMMStudentCont_K3_d5_n1500[[f]]$X, G=K)})
# Robust EM
rgmm <- lapply(1:length(heteroGMMStudentCont_K3_d5_n1500), function(f){
  multipleRobustMM(heteroGMMStudentCont_K3_d5_n1500[[f]]$X, nclust=K)})
```

Again, the effect of contamination on the classification accuracy (in terms of ARI) can be described as follows.

```{r gmmAri, eval=TRUE}
fractionList <- (0:5)/10
plot(fractionList, sapply(1:length(fractionList), function(ff){
  adjustedRandIndex(heteroGMMStudentCont_K3_d5_n1500[[ff]]$Z, gmm[[ff]]$classification)
  }), type='b', pch=20, xlab='fraction of outliers', ylab='ARI')
points(fractionList, sapply(1:length(fractionList), function(ff){
  adjustedRandIndex(heteroGMMStudentCont_K3_d5_n1500[[ff]]$Z, rgmm[[ff]]$bestresult$clusters)
  }), type='b', pch=20, col='red')
```

# References

<div id="refs"></div>
